Stability and accuracy of free surface time integration in viscous flows

### 2016-08-03 free-surface-paper.tex
Make an even number of sign errors...
Add appendix

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
By defining
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^h_i \\
p^h_i
\end{bmatrix} 
\end{aligned}
\label{eq:eigen_substitution}
\end{equation}
and then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
\begin{equation}
(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i.
\label{eq:standard_eigenvalue}
\end{equation}

@@

The implementation of such a scheme results in a slightly 
asymmetric matrix, which can be more difficult to solve, requiring either
changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
An alternative would be to construct a time-integration scheme that remains stable at larger timesteps.
The reason that forward Euler integration performs so badly with the decay of topography is that 
at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
to the \emph{other} side of equilibrium, overshooting even more.
We would like to construct an explicit scheme that accounts for following properties which we know the system has:
\begin{itemize}
\item In the absence of forcing, topography always decreases.
\item Relaxation of small amplitude topography takes the form of exponential decay.
\item The decay mode with the shortest relaxation time is the least stable.
\end{itemize}

@@

>>>>

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It has strictly positive eigenvalues (see Appendix~\ref{app:positive_definite}).
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
By defining
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^h_i \\
p^h_i
\end{bmatrix} 
\end{aligned}
\label{eq:eigen_substitution}
\end{equation}
and then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
\begin{equation}
(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i.
\label{eq:standard_eigenvalue}
\end{equation}

@@

The implementation of such a scheme results in a slightly 
asymmetric matrix, which can be more difficult to solve, requiring either
changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
An alternative would be to construct a time-integration scheme that remains stable at larger timesteps.
The reason that forward Euler integration performs so badly with the decay of topography is that 
at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
to the \emph{other} side of equilibrium, overshooting even more.
We would like to construct an explicit scheme that accounts for following properties which we know the system has:
\begin{itemize}
\item In the absence of forcing, topography always decreases (Appendix~\ref{app:positive_definite}).
\item Relaxation of small amplitude topography takes the form of exponential decay (Section~\ref{sec:eigenvalue}).
\item The decay mode with the shortest relaxation time is the least stable (Section~\ref{sec:timestepping}).
\end{itemize}

@@

\section{The relaxation spectrum is positive definite}
\label{app:positive_definite}
On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
to have only positive eigenvalues $\tau_i$.
An imaginary eigenvalue would have an oscillitory component,
which we do not expect in creeping flows (oscillation being usually due to inertia). 
A negative eigenvalue would grow exponentially, which would violate conservation of energy.
Here we demonstrate the positivity of the relaxation time spectrum.
We could proceed with either the continuous or the discretized version of the governing
equations, but choose the continuous form here.
We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\label{eq:homogeneous_stokes}
\end{equation}
We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
and integrating over the domain:
\begin{equation}
\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
\end{equation}
Integrating by parts and applying boundary conditions yields
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
\end{equation}
Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
Since we are using the incompressible approximation, the second integral is zero.
On a true free surface the boundary stresses are zero, but we make the same approximation
used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
\end{equation}
We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
thereby introducing the relaxation time $\tau_i$:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
\end{equation}
This equation may be solved for $i\tau_i$:
\begin{equation}
\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
\end{equation}
The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
There is the possibility for velocities which do not move the free surface at all, in which case the bottom
integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
relaxation times, these infinite eigenvalues are not usually a problem.
